---
title: "ICP peak in inspiration or in expiration"
subtitle: "Mixed-effects logistic regression on 60-min windows"
author: "Baptiste Balanca"
date: today
lang: en
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    embed-resources: true
    code-fold: show
    code-tools: true
    theme: cosmo
    fig-width: 8
    fig-height: 5
execute:
  warning: false
  message: false
---

## Question and strategy

What determines whether ICP peaks in **inspiration** rather than in **expiration**?

The 60-min windows of a given patient are not independent: a naive test on
11,000 windows from about a hundred patients makes everything significant. We therefore
use a mixed-effects logistic regression (`lme4::glmer`, binomial family) with a random
intercept per patient.

::: {.callout-note}
## Specification choice

**The cluster is not in the model.** The clusters are built from `icp`,
`icp_pulse_amplitude`, `P2P1` and `resp_in_icp`: putting them next to these same variables
amounts to including a deterministic function of its own predictors. The cluster is kept
as a descriptive variable, outside the model.
:::

## Preparation

```{r}
#| label: setup
library(lme4)
library(readxl)
library(dplyr)
library(sjPlot)

# working matrix exported by the last cell of resp_icp_compliance.ipynb
# (filtered windows + ICCA enrichment + cluster label)
path_data <- "df_for_stats/anais_design_matrix_bb_v2.xlsx"
ctrl <- glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
```

```{r}
#| label: data
data <- read_excel(path_data)

# drug doses: an NA means the syringe pump was not running, hence a zero dose.
# Without this, glmer silently drops the affected windows (2556 on sufentanil alone).
drugs <- c("propofol", "sufentanil", "midazolam")

d <- data |>
  filter(label_max_icp_phase %in% c("inspiration", "expiration")) |>
  mutate(
    across(all_of(drugs), \(x) coalesce(x, 0)),
    subject          = factor(subject),
    insp             = as.integer(label_max_icp_phase == "inspiration"),
    ventilation_mode = relevel(factor(ventilation_mode), ref = "controlled"),
    sedation_f       = factor(sedation),
    cluster          = factor(cluster)
  )

print(d)
```

## Model

```{r}
#| label: model
model <- glmer(
  insp ~ icp + icp_pulse_amplitude + P2P1 + resp_rate + ventilation_mode + sedation_f + sufentanil + (1 | subject),
  data = d,
  family = binomial,
  control = ctrl          # bobyqa, defined in the setup cell
)

tab_model(model, show.aic = TRUE, show.icc = TRUE, show.re.var = TRUE, show.obs = TRUE)
```
